---
title: "Working with the KRSP database in R"
author: "Matt Strimas-Mackey"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Working with the KRSP database in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include=F}
knitr::opts_chunk$set(warning = FALSE)
options(knitr.table.format = 'markdown')
```

The `krsp` package offers a set of tools for working with the KRSP MySQL database in R. 
Functions fall into two categories: standard database queries and helper functions for writing custom queries. The former require very little knowledge of R or the database to run. For writing your own queries, this package takes a SQL-free approach, instead using the `dplyr` MySQL interface. Therefore, if you're not already familiar with `dplyr`, it will be useful to read through one of the many [tutorials online](http://seananderson.ca/2014/09/13/dplyr-intro.html) or look through the vignette with `vignette("introduction", "dplyr")`. I provide an extremely brief introduction below.

# `dplyr` basics

`dplyr` is a package designed for manipulating data in data frames quickly and easily. The functions in this package are more consistent, intuitive, and predictable than the equivalent base R functions. There are five core **verbs** (aka functions) for working with single data frames:

- `select()`: select a subset of columns in a data frame.
- `filter()`: select a subset of rows in a data frame.
- `arrange()`: sort the rows in a data frame by the values in a column.
- `mutate()`: add new columns to a data frame that are functions of existing columns.
- `summarise()`: collapse a data frame and summarize data.

Here's a brief demonstration of how these verbs work.

```{r st-verbs}
#install.packages("dplyr")
library(dplyr)
mtc <- tbl_df(mtcars)
mtc <- select(mtc, mpg, cyl, horse_power = hp)
mtc
filter(mtc, cyl == 4, horse_power < 70)
arrange(mtc, desc(cyl), horse_power)
mutate(mtc, mpg_per_hp = mpg / horse_power)
summarize(mtc, mean_mpg = mean(mpg), num_records = n())
```

Note that in all cases the first arguement is a data frame, the subsequent arguements describe what you want to do with the data frame, and the result is always another data frame. Also, columns in the data frame are referenced directly without quotes.

## Chaining operations with `%>%`

The pipe operator, `%>%`, is used to chain multiple operations together in a way that avoids multiple, nested function calls. It can be used with any R functions, but it is particularly useful when working with `dplyr`. Specifically, `x %>% f(y)` is equivalent to `f(x, y)`. Let's look at a simple example, the following two operations are equivalent:

```{r pipe-example}
mean(sqrt(exp(mtc$mpg)))
mtc$mpg %>% 
  exp %>% 
  sqrt %>% 
  mean
```

However, the approach using `%>%` is much easier to read. Let's look at a `dplyr` example:

```{r pipe-dplyr}
mtc %>% 
  filter(cyl == 4, horse_power < 70) %>% 
  mutate(mpg_per_hp = mpg / horse_power) %>% 
  arrange(mpg_per_hp)
```

## Grouped operations

Often you'll want to apply some operation independently to groups of rows in a data frame. The `dplyr` function `group_by()` splits a data frame into groups such that all subsequent operations occur within these groups. `group_by` works with all of the verbs described above, but it's most commonly used with `summarise()`

```{r group-by}
mtc %>% 
  group_by(cyl) %>% 
  summarise(mean_mpg = mean(mpg))
```

## Working with databases

The functions within `dplyr` are usually applied to data frames; however, they can be applied to tables within a SQL database in almost exactly the same way. The only additional steps are to define a connection to the database (i.e. specify where the database is and your login credentials), a process addresssed in the next section. Under the hood, `dplyr` converts function calls to SQL code, but this process is invisible to the user.

One nice feature of `dplyr` is that it tries to be as "lazy" as possible when dealing with databases. This means:

- Data are never pulled from the database unless explicitly asked for, e.g. passed to a function outside `dplyr`.
- The database is only queried at the last minute possible, `dplyr` keeps track of all the intermediate operations and only converts them to SQL and queries the database when results are explicitly requested.
- Only the number of rows explicitly needed are pulled from the database.

This is nice because it means queries to the database and downloading data are kept to an absolute minimum. While `dplyr` documentation calls this behaviour "lazy", I'm inclined to just call it "smart".

# Connecting to the KRSP database

Before accessing tables in a MySQL database you need to create a connection object that defines the location and authentication details for the connection. The function `krsp_connect()` is designed to create this connection object, which you will then pass as a parameter to other functions in this package. I outline three different ways of connecting to the KRSP database for three common scenarios.

## Local instance

The simplest situation is if you have a local copy of the database that you connect to with username `root` and no password. This is the default for `krsp_connect()` so creating a connection object is as simple as:

```r
library(krsp)
con <- krsp_connect()
```

## Remote instance

Things get a little more complicated if you want to connect to a remote KRSP database. 
This could either be a cloud database hosted by Amazon Web Services (AWS) or a database on a computer on your local network, which would be the setup at Squirrel Camp. In either case, you'll need to configure the database correctly so it accepts connections, and set up users, passwords, and privileges. Before you proceed, you'll need the following pieces of information:

- **Host**: the address of the computer where the database lives, either an IP address (e.g. `192.168.0.1`) or a host name (e.g. `krsp.abc123.us-west-2.rds.amazonaws.com`).
- **Username**: a valid username that has been assigned privileges to access the KRSP database.
- **Password**: password for the given username.

With this information there are two ways to connect: the quick/wrong way and the correct/secure way.

### The quick/wrong way

The easiest way to quickly connect to a remote database is to pass all the connection and authentication details directly to `krsp_connect()` as parameters. This is usually a terrible idea because your username and password will be stored in plain text and, in the long run, it will be more work because you'll have to type all the details in each time you connect. Here's how you do it:

```r
con <- krsp_connect(host = "krsp.abc123.us-west-2.rds.amazonaws.com",
                    user = "msm",
                    password = "s0mepaSSword")
```

where the `host`, `user`, and `password` parameters are replaced with the correct values for your situation.

The one situation where this method may make sense is at Squirrel Camp. In this situation the database is isolated from the internet anyway, so it may be easiest to just create a single user (e.g. `squirreler`) with `SELECT` only privileges and no password. Then connecting is as simple as:

```r
con <- krsp_connect(host = "192.168.0.1", user = "squirreler")
```

where `192.168.0.1` should be replaced with the IP address on the local network of the camp computer with the database.

### The correct/secure way

The right way to connect to a remote database, especially in the cloud on AWS, is to specify the connection information in a `my.cnf` connection file. This is a file that MySQL references to retrieve the location of the database and your credentials. This avoids having to store sensitive information, such as your password, within your R code.

On Mac OS and Linux the connection file is `~/.my.cnf` and on Windows it is located at `C:/my.cnf`. Open this file, or create it if it doesn't already exist, and enter the following text:

```
[krsp-aws]
username=msm
password=s0mepaSSword
host=krsp.abc123.us-west-2.rds.amazonaws.com
port=3306
database=krsp
```

Make sure you edit this to include your own username and password, and the correct host, which is the Public DNS of the AWS instance or the IP address of a remote computer. Multiple connection profiles can be set up with different names, here I've used `krsp-aws` for the name of this profile.

To reduce the risk of someone getting access to your database credentials, it's wise to set the permissions of the `.my.cnf` file so only you can read it. On Mac OS or Linux, open the Terminal and enter the following command:

```bash
sudo chmod 600 ~/.my.cnf
```

To create a connection object provide the name of the `my.cnf` profile to the `group` parameter of `krsp_connect()`:
 
```r
con <- krsp_connect(group = "krsp-aws")
```
